Best Practices for Alchemical Free Energy Calculations: v\versionnumber

### 2019-12-09 paper/manuscript.tex
Updates from Overleaf

%\label{sec:abbreviations}
\begin{tcolorbox}[title=Abbreviations, colback=blue!10!white]
    \textit{CPU} - Central Processing Unit\\
    \textit{FE} - Free Energy\\
    \textit{FEP} - Free Energy Perturbation\\
    \textit{GPCR} - G-Protein Coupled Receptor\\
    \textit{GPU} - Graphics Processing Unit\\
    \textit{MBAR} - Multistate Bennett Acceptance Ratio\\
    \textit{MCSS} - Maximum Common SubStructure\\
    \textit{MD} - Molecular Dynamics\\
    \textit{RBFE} - Relative Binding Free Energy\\
    \textit{RMSE} - Root Mean Square Error\\
    \textit{SAR} - Structure-Activity Relationships\\
    \textit{TI} - Thermodynamic Integration
\end{tcolorbox}

\section{What are alchemical free energy methods?}
\label{sec:intro}
Alchemical free energy (FE) calculations compute free energy differences associated with transfer processes, such as the binding of a small molecule to a receptor or the transfer of a small molecule from an aqeuous to apolar phase~\cite{zwanzig1954hightemperature}. 
The calculations use non-physical intermediate states where atoms are changed from one to another or can be removed or inserted by decoupling or coupling with their environment, Figure~\ref{fig:fig1_what_is_alchemy}. These so-called alchemical states permit converged calculations for large changes, such as the entire removal of a ligand from a binding site, known as an absolute FE calculation, or modifying a substituent between two analogous ligands in a binding site, a relative FE calculation. Two common approaches are free energy perturbation (FEP) and thermodynamic integration (TI) for which the theory dates back decades but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamics, merz1989free}. Important studies continued during the 2000's often leading to improved implementation of the methods in common simulation software~\cite{vanderspoel2005gromacs, mermelstein2018fast, wang2015accurate, biosimspace}. This foundation work, combined with the methodological, technological and hardware improvements of the last 5-10 years have led to an explosion of interest.  The domain of applicability for these calculations now involves such varied applications as the computation of protein-ligand binding free energies (both relative~\cite{relative-binding-free-energies} and absolute~\cite{absolute-binding-free-energies}), ligand selectivities~\cite{selectivity}, partition and distribution coefficients between different liquid phases (such as octanol-water partition coefficients~\cite{octanol-water-partition}), affinity changes due to mutations in therapeutic targets~\cite{hauser2018predicting,aldeghi2018accurate}, and changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate}. As the field of molecular simulation moves towards millisecond timescales accurate FE calculations on even more challenging problems will be in scope. In the meantime, today's user may find it hard to get started with these complex calculations whilst also keeping up with the fast pace of change. This best practice guide aims to serve as a go-to reference for current recommendations and tips for users of all experience.

@@

Below, we provide a non-exhaustive list of topics we have not addressed with some references to provide starting points on these more advanced topics:
\begin{itemize}
\item Covalent inhibition.
\item Complex association stoichiometry (beyond simple 1:1 association)~\cite{awesome reference}. 
\item Endpoint free energy methods such as MM-PBSA~\cite{genheden2015mm} and LIE~\cite{gutierrez-de-terran2012linear}
\item Free energy methods that extract the ligand using geometric order parameters and potential of mean force methods~\cite{heinzelmann2017attachpullrelease}
\item Forcefield dependence for protein, ligand, ions, co-solvents, and co-factors. Different studies have looked at the influence of force fields and it is assumed the user has made an adequate choice for the system under study.~\cite{loeffler2018reproducibility, vassetti2019assessment, lopes2015current} 
\end{itemize}

@@

The concerns mentioned in the previous section such as understanding if the series of molecules is suitable, the simulation timescales relevant to capture the structure activity relationships (SAR), and performing up-front tests of performance are all relevant to drug discovery applications. However, without venturing too far into details of system setup, which is beyond the scope of this article, there are some additional points to consider. 
\subsection*{Capturing the experimental conditions?}
The calculations aim to capture the alchemical change from one ligand to another as accurately as possible. Therefore, it is necessary to consider details of the experimental setup, such as pH. AssaysFormal charge, or tautomeric state of the small molecules can change within a series. The pKa can change by several log units with only small structural modifications and this is a tactic often used within a congeneric series to modify drug properties. Changes in charge or tautomers would require modified approaches. A second concern is to try to capture aspects of the bioassay into the simulations, for instance what pH was used for screening? Does the protein X-ray match the construct used for screening (i.e. catalytic vs full length, monomer vs dimer) (cite Perez-Benito Sci Rep 2018, 8, 4883)? Were certain co-factors or partner proteins required to augment activity? 
%
\subsection*{Is my binding mode accurate?}
As also mentioned, good performance of alchemical FE calculations requires an accurate representation of the ligand binding mode, normally solved via X-ray crystallography. If more than one structure is available, the modeler should pay attention to choose the most suitable. The quality of the structure can be a concern, and the reader is referred to work of Warren et al for a detailed discussion of choosing optimal structures for structure-based modeling (CITE: Warren et al Drug Discov Today, 2012, 17, 1270). Particularly relevant for alchemical FE calculations is to study the SAR and understand the impact in the binding site, is side chain movement required and is there evidence of this in any alternative X-ray structures of the same protein? Normally, only one protein and water configuration is used, so this needs to be capable of accommodating the smallest through to largest ligands in a way that allows stable and well behaved simulations. This can be a practical limitation for the extent of structural change that can be attempted within a series, a simple work around can be to separate compounds into subseries for different calculations. If multiple structures are available there is some evidence the higher affinity complex can give better performance (cite Perez-Benito jctc 2019, 15, 1884). 
%

In a drug discovery setting it is normal to consider dozens (or more) of ligands and how to align and place them in the binding site has to be defined. There is no detailed study of how different approaches to this may affect results, but the user should be aware of some practical considerations. Tools are available to compare the ligands and build the hybrid topologies that define the changes between one ligand and another. In simple terms, providing poor alignment to these tolls will make this job harder. Therefore, docking ligands with no restraints is not usually the best method for defining the ligand inputs, instead a core restrained docking can provide satisfactory outcomes, but in this case one still should pay careful attention to consistency of alignment for identical substituents. Another alternative often used is to manually edit the same core with the changing substituents. This provides assurances that coordinates for the non-perturbed structures remain identical and aromatic substituents for instance obey consistent dihedral angles. This method is of course prohibitive for many compounds so automating this approach can be desirable. Finally, the role of water in ligand binding is now well understood and it is a crucial aspect to accurately reflect the changes in binding site solvation during ligand binding. Can crystallographic waters be retained? Do they clash with some of the larger ligands used in the FEP? If so, how will the water be incorporated for smaller ligands that require the water – consider pre-solvation methods such as GCMC steps (CITE Julien michel J. Essex paper?). Tools such as watermap or opensource equivalents can be used to help define water structure for systems with no experimental evidence of water sites.

Running calculations in a drug discovery application will typically require the use of software or tools to facilitate the calculations of many compounds. The well-known commercial implementation from Schrodinger, FEP+, uses a graphical user interface and once the user has calculated any missing force field parameters, calculations can be prepared and submitted within minutes. The software takes care of building the map of perturbations, connecting similar compounds based on LOMAP methodology (CITE LOMAP), all inputs for the Desmond simulations are built automatically and the job can be saved or run immediately on available GPUs. The cost of this software is often prohibitive but alternative tools are becoming available to allow relatively straight forward running of alchemical FE calculations with other software (CITE New Merz paper, cite PMX paper Vytas and Bert de Groot, cite FESETUP, cite FEanalysis). One consideration at this point is to define the perturbations suitable for the application. For instance, if the goal is to accurately assess the relative binding energy of a small number of compounds, possibly with challenging synthesis, the map of perturbations may contain as many connections between compounds as affordable. However, when running calculations on hundreds of compounds a so called ‘star-map’ can be used that just contains one connection per compound: perturbing every compound to a central ligand, typically the crystal structure ligand. In this way the top-ranking examples can be readily identified and submitted to additional calculations in a second round. 
%

For prospective drug discovery applications there are several other considerations including understanding likely errors and taking selection bias into account. It is crucial when proposing compounds for synthesis to have some idea of the underlying error or uncertainty in the predictions. Fortunately, there are several useful indicators, results suggest MBAR errors for perturbations greater than 0.3 kcal/mol can be indicative of inaccuracy (cite Perez-Benito jctc 2019, 15, 1884). In addition, the hysteresis can be checked by comparing the difference for forward and reverse perturbations of the same structural change, or in the case of large maps of perturbations cycle closure hysteresis (cite Wang et al JCTC 2013, 9, 1282) indicates problematic perturbations involved in cycles connecting many compounds. Selection bias, as described above, can be problematic in prospective applications as there is normally an aim to focus on only a narrow range of predicted activity (i.e. the most active). Thus, correlation between prediction and subsequent experiment may lead to disappointment but corrections can be applied based on previous recommendations (Abel et al Curr Top med Chem 2017, 17,2577). 
%

In summary, the key aspects for successful drug discovery application are to work within the domain of applicability using high quality X-ray structure of the target compounds and testing the approach with a retrospective validation study to ensure the basics of system setup are understood and under control. Always assess the confidence in the predictions and communicate this when discussing with experimentalists. Consider performing repeat calculations for at least some of the perturbations in the study. In a drug discovery setting it is inevitable to consider larger virtual libraries, in these cases again evaluate if shorter simulation time per $\lambda$ window can be sufficient for a ‘virtual screening’ mode of application. It is expected that future tools may allow efficient exploration of large virtual libraries. There are many accounts of success of alchemical FE calculations, the methods are undoubtedly showing good performance towards the goal of binding energy prediction. However, it is important to have realistic expectations. Structure based drug design projects are often capable of improving potency relatively quickly and the range of activity narrows to just two-to-three log units. It may seem hard to impact with substantially different, more potent, stand-out compounds in this scenario, but binding energy prediction can still be extremely useful for ensuring activity is maintained as other properties are optimized. An interesting cost benefit analysis has been performed, see drug discovery strategy related discussions of Mobley Abel and co. From a drug discovery point of view, FEP is expanding its domain of applicability, there are reports of success using homology models and GPCRs for instance, as well as enabling charge change and scaffold hopping, but these systems are undoubtedly more difficult.  In the meantime the use cases are expanding to resistance prediction, disease mutation prediction, solubility prediction – an exciting future for alchemical FE calculations.

%
%
%
%%%%%%%%%%%%
% Step 1   %
%%%%%%%%%%%%
%
\section{Simulation prerequisites}
\label{sec:step1}
\todo[inline,color={green!20}]{@everyone: please review}
The alchemical free energy protocols, which are discussed in Section~\ref{sec:step2}, which can be used are tied in with what type of free energy you might want to compute, i.e. a free energy of binding or a free energy of hydration. Depending on the type of simulations different considerations have to be made for ligands, solvent, and host molecules (in the case of the estimation of free energies of binding).
\subsection*{Free energies of binding}
In principle in the limit of sufficient conformational sampling, the free energy changes estimated from an alchemical free energy calculation should be independent of the system's initial coordinates. However, in practice because simulations of thermodynamic states are of finite duration (typically currently 1-100 ns per state) this is only true for certain classes of alchemical free energy calculations such as relative or absolute free energies of hydration of small and relatively rigid organic molecules. Host-guest or protein-ligand complexes typically exhibit slowly relaxing degrees of freedom that exceed significantly the duration of an alchemical free energy calculation. It is therefore generally important to carefully select input coordinates to obtain satisfactory results. 
You might want to ask yourself the following questions before diving into your simulation setup. 
\begin{itemize}
    \item Do I have one or multiple good host structures? (e.g. a good resolution X-ray crystal of the protein target)
    \item Should I include burried waters, or other small molecules that can be found in an X-ray structure.
    \item Are my ligands part of a congeneric series? (simple R group substitutions around the same scaffold)
    \item Do I have information on one or all of the ligands binding sites (e.g. a X-ray structure)
\end{itemize}

As for any simulation, good care should be taken in selecting protein structures by looking at available X-ray structures in the protein database and potentially selecting multiple structures initially, to account for variability in receptor conformations as well as accuracy of available X-ray structures. Typically, clustering of receptor structures can be used to identify different receptor conformations near the binding site, as well as assessing relevant side chain placements from the X-ray structure, see for example~\cite{mey etal, the other D3R paper I can't remember}. In terms of set up and other choices following general best practice guidelines is advisable~\cite{braun2018best}.

For binding free energy calculations using a congeneric set of ligands and therefore likely to select a relative free energy protocol (see~\ref{sec:step2}), other than the choice of forcefield for the parametrization of the ligands, some care has to be taken selecting binding poses for these ligands. Generally a common assumption for congeneric is that the binding mode is conserved within the series. Therefore if an X-ray structure of one of the ligands is available, this should be used to position the congeneric series in its putative binding site in an energetically reasonable conformation that entails no steric or electrostatic mismatch with the receptor. Get in the habit of looking at X-ray structures and their electron densities properly, as often the resolution of part of the ligand may be more based on the interpretation of the crystallographer than the available electron density. For example, looking at a cyclo-hexane ring density, a chair conformation is vastly more likely than that of a boat~\cite{maybe we can find some reference here?} and may just be poorly assigned.

@@

Congeneric series can contain sterioismoerms or enantiomers. Making sure that you model the correct enantiomers or sterioisomers is important.
\todo[inline, color=green!20]{Question: What is best practice here? I tend to pick one enantiomer and just model that, but experimentally they may have used a racemic mixture?}

Another aspect to look out for are binding site water molecules that may form water mediated protein-ligand interactions, particularly in situations where exchange with bulk water may be slow on the simulation timescales. This happens typically in buried binding site. If multiple protein X-ray structures are available, retaining conserved water molecules is an easy way to not miss waters. In cases where water molecules are known to play an important role in the binding, software implementations that use water sampling facilitated by Grand Canonical Monte Carlo methods may be a good way to go. For examples see \textcolor{red}{system1, system2, and system3}~\cite{x,y,z}.

@@

If the binding site location is uncertain but the structure of the receptor is well defined and plausible binding sites are identified, you are more likely to chose an absolute free energy protocol to compute the standard free energy of binding of the ligand to a set of binding sites. This requires the user to prepare input files describing the bound conformation in different putative binding sites (REF). The apparent binding free energy of the ligand may be obtained by combining the individual binding site free energies, which also indicate where the ligand is more likely to bind. In this case using a docking program to generate initial structures is the way to go. Different commercial and none commercial tools are available, such as: rDock, Vina, glide, or Flare to name a few~\cite{rDock, Vina, Glide, Flare}. 
If the putative binding sites are not apparent because of for instance significant induced-fit effects it may be challenging to obtain meaningful free energies of binding. One would have to account for the free energy change for forming a binding site in the target receptor. However, it may still be possible to obtain useful binding free energy estimates that may be compared between different ligands.

\subsection*{Free energies of hydration or partition coefficients}
Preliminary considerations necessary for using free energy methods to compute partition coefficients are much generally much more straight forward. Using even something such as bable to generate a 3D minimised structure of the solute you would like to compute a free energy of hydration or partition coefficient can be sufficient for preparation of these types of simulations. However, in these cases a careful choice of forcefields, as well as water models or organic solvents is essential. See for example XX.et al~\cite{xxx} for a good discussion on these choices. 
\todo[inline, color=green!20]{Just a note as a reminder: Chirality!}
%
%
%
%%%%%%%%%%%%
% Step 1   %
%%%%%%%%%%%%
%
\section{What simulation protocol should I choose? -- Step 2}
\label{sec:step2}
Alchemical free energy calculations can be grouped into two main categories, ``absolute'' and ``relative'' \footnote{The distinction is a bit of a misnomer, since both compute ratios of partition functions relative to another state, and neither computes an absolute free energy.}, which differ in whether they compute properties for a single molecule (absolute) or compare properties of different, usually closely related, molecules (relative).
To use binding as a concrete example, in absolute binding free energy calculations, one computes the binding free energy of a ligand to an individual receptor relative to a standard reference concentration.
In contrast, in relative binding free energy calculations, one compares the binding free energy of two related inhibitors to determine the potency difference.
\subsection*{Absolute and relative free energy calculations have some differences}
Many protocol issues for alchemical calculations are common, but some are different between absolute and relative calculations, so before treating the common elements we treat the protocol differences.

@@

There are different tools that allow the generation of MCSS matches as well as single topology input. A large number of software tools are nowadays available for computing MCSS matches, some of the rely on RDKit for this computation, such as LOMAP or FESetup and partially BioSimSpace, others such as fkckombu are standalone tools~\cite{rdkit, lomap, fesetup, biosimspace, fkcombu}. Schr\"{o}dinger's FEP+ planning tool is based on a version of LOMAP, also allows MCSS matching and planning single topology calculations between molecules~\cite{FEP+}.

@@

\paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic, in part because relative calculations rely on the free energy contributions of dummy atoms canceling between different legs of the thermodynamic cycle [refs], which may not be true whenever dummy atoms are involved in rings.
Some approaches have attempted to address this [ref Schrodinger] but a general solution is not yet in mainstream use.

>>>>

%\label{sec:abbreviations}
\begin{tcolorbox}[title=Acronyms, colback=blue!10!white]
    \textit{CPU} - Central Processing Unit\\
    \textit{FE} - Free Energy\\
    \textit{FEP} - Free Energy Perturbation\\
    \textit{GPCR} - G-Protein Coupled Receptor\\
    \textit{GPU} - Graphics Processing Unit\\
    \textit{MBAR} - Multistate Bennett Acceptance Ratio\\
    \textit{MCSS} - Maximum Common SubStructure\\
    \textit{MD} - Molecular Dynamics\\
    \textit{RBFE} - Relative Binding Free Energy\\
    \textit{RMSE} - Root Mean Square Error\\
    \textit{SAR} - Structure-Activity Relationships\\
    \textit{TI} - Thermodynamic Integration
\end{tcolorbox}
%\label{sec:abbreviations}
\begin{tcolorbox}[title=List of Symbols, colback=green!10!white]
U - Potential energy
\end{tcolorbox}

\section{What are alchemical free energy methods?}
\label{sec:intro}
Alchemical free energy (FE) calculations compute free energy differences associated with transfer processes, such as the binding of a small molecule to a receptor or the transfer of a small molecule from an aqeuous to apolar phase~\cite{zwanzig1954hightemperature}. 
The calculations use non-physical intermediate states where atoms are changed from one to another or can be removed or inserted by decoupling or coupling with their environment, Figure~\ref{fig:fig1_what_is_alchemy}. These so-called alchemical states permit converged calculations for large changes, such as the entire removal of a ligand from a binding site, known as an absolute FE calculation, or modifying a substituent between two analogous ligands in a binding site, a relative FE calculation. Two common approaches are free energy perturbation (FEP) and thermodynamic integration (TI) for which the theory dates back decades but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamics, merz1989free}. Important studies continued during the 2000's often leading to improved implementation of the methods in common simulation software~\cite{vanderspoel2005gromacs, mermelstein2018fast, wang2015accurate, hedges2019biosimspace}. This foundation work, combined with the methodological, technological and hardware improvements of the last 5-10 years have led to an explosion of interest.  The domain of applicability for these calculations now involves such varied applications as the computation of protein-ligand binding free energies (both relative~\cite{relative-binding-free-energies} and absolute~\cite{absolute-binding-free-energies}), ligand selectivities~\cite{selectivity}, partition and distribution coefficients between different liquid phases (such as octanol-water partition coefficients~\cite{octanol-water-partition}), affinity changes due to mutations in therapeutic targets~\cite{hauser2018predicting,aldeghi2018accurate}, and changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate}. As the field of molecular simulation moves towards millisecond timescales accurate FE calculations on even more challenging problems will be in scope. In the meantime, today's user may find it hard to get started with these complex calculations whilst also keeping up with the fast pace of change. This best practice guide aims to serve as a go-to reference for current recommendations and tips for users of all experience.

@@

Below, we provide a non-exhaustive list of topics we have not addressed with some references to provide starting points on these more advanced topics:
\begin{itemize}
\item Covalent inhibition.
\item Complex association stoichiometry (beyond simple 1:1 association)~\cite{awesome reference}. 
\item Endpoint free energy methods such as MM-PBSA~\cite{genheden2015mm} and LIE~\cite{utierrez-de-teran2012lineara}
\item Free energy methods that extract the ligand using geometric order parameters and potential of mean force methods~\cite{heinzelmann2017attachpullrelease}
\item Forcefield dependence for protein, ligand, ions, co-solvents, and co-factors. Different studies have looked at the influence of force fields and it is assumed the user has made an adequate choice for the system under study.~\cite{loeffler2018reproducibility, vassetti2019assessment, lopes2015current} 
\end{itemize}

@@

The concerns mentioned in the previous section such as understanding if the series of molecules is suitable, the simulation timescales relevant to capture the structure activity relationships (SAR), and performing up-front tests of performance are all relevant to drug discovery applications. However, without venturing too far into details of system setup, which is beyond the scope of this article, there are some additional points to consider. 
\subsection*{Capturing the experimental conditions}
The calculations aim to capture the alchemical change from one ligand to another as accurately as possible. Therefore, it is necessary to consider details of the experimental setup, such as pH. Biological assays are usually run at neutral pH but is not always the case, and protein and ligand preparation protocols should reflect this. In more detail, the formal charge, or tautomeric state of the small molecules can change within a series of analogs. The change of pKa is a common medicinal chemistry strategy to modify drug properties and would normally require explicit effort to incorporate into alchemical calculations. Further concerns about the experimental setup require understanding what protein construct is used in the bioassay. For instance, does the X-ray that is to be used for the calculations match the construct used for screening (i.e. catalytic vs full length, monomer vs dimer) (cite Perez-Benito Sci Rep 2018, 8, 4883)? Also, were certain co-factors or partner proteins required to augment activity? 
\subsection*{Is my binding mode accurate?}
As also mentioned, good performance of alchemical FE calculations requires an accurate representation of the ligand binding mode, normally solved via X-ray crystallography. If more than one structure is available, the modeler should pay attention to choose the most suitable. The quality of the structure can be a concern, and the reader is referred to work of Warren et al. for a detailed discussion of choosing optimal structures for structure-based modeling~\cite{warren2012essential}. It is also useful to study the structure activity relationship and understand the expected impact in the binding site, is side chain movement required and is there evidence of this in any alternative X-ray structures of the same protein? Normally, only one protein and water configuration is used for a series of alchemical FE calculations, so this needs to be capable of accommodating the smallest through to largest ligands in a way that allows stable and well behaved simulations. This can be a practical limitation for the alchemical changes that are feasible, a simple work around can be to separate compounds into sub-series for different calculations. If multiple structures are available there is some evidence the higher affinity complex can give better performance~\cite{perez-benito2019predicting}.
\subsection*{Input setup and scale of calculations}
In a drug discovery setting it is normal to consider dozens (or more) of ligands and it is necessary to align them in the binding site. There is no detailed study of how different alignment approaches may affect results, but the user should be aware of some practical considerations. Tools are available to compare the ligands and build the hybrid topologies that define the changes between one ligand and another~\cite{loeffler2015fesetup,hedges2019biosimspace,gapsys2015pmx}. In simple terms, providing poor alignment to these tools will make this job harder. Therefore, docking ligands with no restraints is not usually the best method for defining the ligand inputs, instead a core restrained docking can provide satisfactory outcomes, but in this case one still should pay careful attention to consistency of alignment for identical substituents. Another alternative often used is to manually edit the same core with the changing substituents. This provides assurances that coordinates for the non-perturbed structures remain identical and aromatic substituents for instance obey consistent dihedral angles. This approach is not feasible for many compounds and therefore automating this approach can be desirable. Finally, the role of water in ligand binding is now well understood and it is a crucial aspect to accurately reflect the changes in binding site solvation during ligand binding. Can crystallographic waters be retained? Do they clash with some of the larger ligands used in the FEP? If so, how will the water be incorporated for smaller ligands that require the water – consider pre-solvation methods such as GCMC steps~\cite{michel2010prediction}. Tools such as watermap or opensource equivalents (SSTmap, others) can be used to help define water structure for systems with no experimental evidence of water sites. Before launching large numbers of alchemical free energy calculations it is recommended to test the defined system using classical MD simulations and limited numbers of alchemical perturbations. Metrics such as ligand and protein RMSD and RMSF can be inspected, along with visual inspection of simulations, to ensure the system is stable and likely to be suitable for FE calculations. 
Running binding FE calculations in a drug discovery application will typically require the use of software or tools to facilitate the calculations of many compounds. The commercial implementation from Schrodinger, FEP+, uses a graphical user interface and once the user has calculated any missing force field parameters, calculations can be prepared and submitted within minutes. The software builds the network of perturbations, connecting similar compounds based on LOMAP methodology~\cite{liu2013lead}, all inputs for the Desmond simulations are built automatically and the job can be saved or run immediately on available GPUs. The cost of this software is often prohibitive but alternative tools are becoming available to allow relatively straight forward running of alchemical FE calculations with other software~\cite{gapsys2015pmx, loeffler2015fesetup} (CITE New Merz paper, cite PMX paper Vytas and Bert de Groot, cite FESETUP, cite FEanalysis). One consideration at this point is to define the perturbations suitable for the application. For instance, if the goal is to accurately assess the relative binding energy of a small number of compounds, possibly with challenging synthesis, the map of perturbations may contain as many connections between compounds as affordable. However, when running calculations on hundreds of compounds a so called ‘star-map’ can be used that just contains one connection per compound: perturbing every compound to a central ligand, typically the crystal structure ligand. In this way the top-ranking examples can be readily identified and submitted to additional calculations in a second round. 
\subsection*{Making predictions, understanding errors}
For prospective drug discovery applications there are several other considerations including understanding likely errors and taking selection bias into account. It is crucial when proposing compounds for synthesis to have some idea of the underlying error or uncertainty in the predictions. Fortunately, there are several useful indicators, results suggest MBAR errors for perturbations greater than 0.3 kcal/mol can be indicative of inaccuracy~\cite{perez-benito2019predicting}. In addition, the hysteresis can be checked by comparing the difference for forward and reverse perturbations of the same structural change, or in the case of large maps of perturbations cycle closure hysteresis~\cite{wang2013modeling} indicates problematic perturbations involved in cycles connecting many compounds. Selection bias, as described above, can be problematic in prospective applications as there is normally an aim to focus on only a narrow range of predicted activity (i.e. the most active). Thus, correlation between prediction and subsequent experiment may lead to disappointment but corrections can be applied based on previous recommendations~\cite{abel2017critical}. 

In summary, the successful use of alchemical calculations in drug discovery requires working in the domain of applicability, using a high quality X-ray structure of the target compounds, and testing the approach retrospectively to ensure the system setup is well-behaved. Always assess the confidence in the predictions and communicate this when discussing with experimentalists. Consider performing repeat calculations for at least some of the perturbations in the study. In a drug discovery setting it is inevitable to consider larger virtual libraries, in these cases again evaluate if shorter simulation time per $\lambda$ window can be sufficient for a ‘virtual screening’ mode of application. It is expected that future tools may allow efficient exploration of large virtual libraries. There are many accounts of success of alchemical FE calculations, the methods show good performance towards the goal of binding energy prediction. However, it is important to have realistic expectations. Structure based drug design projects are often capable of improving potency relatively quickly and the range of activity narrows to just two-to-three log units. It may seem hard to impact with substantially different, more potent, stand-out compounds in this scenario, but binding energy prediction can still be extremely useful for ensuring activity is maintained as other properties are optimized. An interesting cost benefit analysis has been performed, see drug discovery strategy related discussions of Mobley Abel and co. From a drug discovery point of view, FEP is expanding its domain of applicability, there are reports of success using homology models and GPCRs for instance, as well as enabling charge change and scaffold hopping, but these systems are undoubtedly more difficult.  In the meantime the use cases are expanding to resistance prediction, disease mutation prediction, solubility prediction – an exciting future for alchemical FE calculations.

%
%
%
%%%%%%%%%%%%
% Step 1   %
%%%%%%%%%%%%
%
\section{Simulation prerequisites}
\label{sec:step1}
\todo[inline,color={green!20}]{@everyone: please review}
The alchemical free energy protocols, that are discussed in Section~\ref{sec:step2}, are defined for a specific type of free energy calculation, i.e. a free energy of binding or a free energy of hydration. Depending on the type of simulations different considerations have to be made for ligands, solvent, and host molecules (in the case of the estimation of free energies of binding).
\subsection*{Free energies of binding}
In principle in the limit of sufficient conformational sampling, the free energy changes estimated from an alchemical free energy calculation should be independent of the system's initial coordinates. However, in practice because simulations of thermodynamic states are of finite duration (typically currently 1-100 ns per state) this is only true for certain classes of alchemical free energy calculations such as relative or absolute free energies of hydration of small and relatively rigid organic molecules. Host-guest or protein-ligand complexes typically exhibit slowly relaxing degrees of freedom that exceed significantly the duration of an alchemical free energy calculation. It is therefore generally important to carefully select input coordinates to obtain satisfactory results. 
You might want to ask yourself the following questions before diving into your simulation setup. 
\begin{itemize}
    \item Do I have one or multiple good host structures? (e.g. a good resolution X-ray crystal of the protein target)
    \item Should I include burried waters, or other small molecules that can be found in an X-ray structure.
    \item Are my ligands part of a congeneric series? (simple R group substitutions around the same scaffold)
    \item Do I have information on one or all of the ligands binding sites (e.g. a X-ray structure)
\end{itemize}

As for any simulation, good care should be taken in selecting protein structures by looking at available X-ray structures in the protein database and potentially selecting multiple structures initially, to account for variability in receptor conformations as well as accuracy of available X-ray structures. Typically, clustering of receptor structures can be used to identify different receptor conformations near the binding site, as well as assessing relevant side chain placements from the X-ray structure, see for example~\cite{mey2016blinded}. In terms of set up and other choices following general best practice guidelines is advisable~\cite{braun2018best}.

For binding free energy calculations using a congeneric set of ligands and therefore likely to select a relative free energy protocol (see~\ref{sec:step2}), other than the choice of forcefield for the parameterization of the ligands, some care has to be taken selecting binding poses for these ligands. Generally a common assumption for congeneric is that the binding mode is conserved within the series. Therefore if an X-ray structure of one of the ligands is available, this should be used to position the congeneric series in its putative binding site in an energetically reasonable conformation that entails no steric or electrostatic mismatch with the receptor. Get in the habit of looking at X-ray structures and their electron densities properly, as often the resolution of part of the ligand may be more based on the interpretation of the crystallographer than the available electron density. For example, looking at a cyclo-hexane ring density, a chair conformation is vastly more likely than that of a boat and may just be poorly assigned.

@@

Congeneric series can contain stereoisomers or enantiomers, making sure that you model the correct one is important.
\todo[inline, color=green!20]{Question: What is best practice here? I tend to pick one enantiomer and just model that, but experimentally they may have used a racemic mixture?}

Another aspect to look out for are binding site water molecules that may form water mediated protein-ligand interactions, particularly in situations where exchange with bulk water may be slow on the simulation timescales. This happens typically in buried binding site. If multiple protein X-ray structures are available, retaining conserved water molecules is an easy way to not miss waters. In cases where water molecules are known to play an important role in the binding, software implementations that use water sampling facilitated by Grand Canonical Monte Carlo methods may be a good way to go. For examples see HSP90 as formed part of the D3R grand challenge 2015~\cite{mey2016blinded},A2A~\cite{brucemacdonald2018ligand}, MUP~\cite{ross2015water}, and others~\cite{michel2009energetics}.

@@

If the binding site location is uncertain but the structure of the receptor is well defined and plausible binding sites are identified, you are more likely to chose an absolute free energy protocol to compute the standard free energy of binding of the ligand to a set of binding sites. This requires the user to prepare input files describing the bound conformation in different putative binding sites (REF). The apparent binding free energy of the ligand may be obtained by combining the individual binding site free energies, which also indicate where the ligand is more likely to bind. In this case using a docking program to generate initial structures is the way to go. Different commercial and none commercial tools are available, such as: rDock, Autodock Vina, Glide, or Flare to name a few~\cite{ruiz-carmona2014rdock, trott2010autodock, friesner2004glide, Flare}. 
If the putative binding sites are not apparent because of for instance significant induced-fit effects it may be challenging to obtain meaningful free energies of binding. One would have to account for the free energy change for forming a binding site in the target receptor. However, it may still be possible to obtain useful binding free energy estimates that may be compared between different ligands.

\subsection*{Free energies of hydration or partition coefficients}
Preliminary considerations necessary for using free energy methods to compute partition coefficients are much generally much more straight forward. Using even something such as bable to generate a 3D minimised structure of the solute you would like to compute a free energy of hydration or partition coefficient can be sufficient for preparation of these types of simulations. However, in these cases a careful choice of forcefields, as well as water models or organic solvents is essential. See for example~\cite{bosisio2016blinded,rustenburg2016measuring} for a good discussion on these choices. 
\todo[inline, color=green!20]{Just a note as a reminder: Chirality!}
%
%
%
%%%%%%%%%%%%
% Step 1   %
%%%%%%%%%%%%
%
\section{What simulation protocol should I choose? -- Step 2}
\label{sec:step2}
Alchemical free energy calculations can be grouped into two main categories, ``absolute'' and ``relative'' \footnote{The distinction is a bit of a misnomer, since both compute ratios of partition functions relative to another state, and neither computes an absolute free energy.}, which differ in whether they compute properties for a single molecule (absolute) or compare properties of different, usually closely related, molecules (relative).
To use binding as a concrete example, in absolute binding free energy calculations, one computes the binding free energy of a ligand to an individual receptor relative to a standard reference concentration.
In contrast, in relative binding free energy calculations, one compares the binding free energy of two related inhibitors to determine the potency difference.
\subsection*{Absolute and relative free energy calculations have some differences}
Many protocol issues for alchemical calculations are common, but some are different between absolute and relative calculations, so before treating the common elements we treat the protocol differences.

@@

There are different tools that allow the generation of MCSS matches as well as single topology input. A large number of software tools are nowadays available for computing MCSS matches, some of the rely on RDKit for this computation, such as LOMAP or FESetup and partially BioSimSpace, others such as fkckombu are standalone tools~\cite{rdkit,loeffler2015fesetup, liu2013lead, hedges2019biosimspace, kawabata20143d}. Schr\"{o}dinger's FEP+ planning tool is based on a version of LOMAP, also allows MCSS matching and planning single topology calculations between molecules~\cite{wang2015accurate}.

@@

\paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic, in part because relative calculations rely on the free energy contributions of dummy atoms canceling between different legs of the thermodynamic cycle~\cite{}, which may not be true whenever dummy atoms are involved in rings.
Some approaches have attempted to address this [ref Schrodinger] but a general solution is not yet in mainstream use.

\paragraph{Perturbation maps}
As part of LOMAP and the FEP+ planning tool. Based on the input ligand series a perturbation map can be planned. Recent heuristics have shown the more connected the perturbation network the better, however there is away to optimise network structure while minimising the number of perturbations that need to be computed minimising the resulting computational cost~\cite{yang2019optimal}. Sometimes the introduction of intermediates that are not part of the original congeneric series are essential to avoid ring breaking, or dealing with perturbations that would otherwise result in large numbers of atoms to be grown or disappearing. None commercial tools have underlying good heuristic but may fail with complicated input, needing user validation in particular when dealing with chiral compounds.
